Ion pairing controls rheological properties of ``crocodile-line'' polyelectrolyte hydrogels

### 2016-01-29 ion_pairing_paper.tex
weight fraction vs molar concentration paves the way to supra-molecular assembly

Between two cross-links the crocodile chain can be considered as an entropic spring made of $N=n N_0$ monomers. The shear modulus of the gel is then given by\cite{Rubinstein1996}:
\begin{equation}
G = \frac{c}{N}k_\mathrm{B}T,
\label{eq:G}
\end{equation}
where $c$ is the monomer number density, $k_\mathrm{B}$ is the Boltzmann constant and $T$ the temperature. Knowing the molecular mass $M$ of the chains, density $d$ of the solvent and the weight fraction $w$ of polymer, we obtain
\begin{equation}
n = \frac{\mathcal{N_A}}{M} w d \frac{k_\mathrm{B}T}{G}.
\end{equation}

We find that the number $n$ of chains between cross-link point goes from 1 in \ce{PPyr+I-} and \ce{PPy+Br-} to 790 in \ce{PIm+Br-}, following the \textit{a priori} ranking of charge dissociation, see Table~\ref{tab:results}. This is coherent with a higher probability of attaching two or more heads on a highly dissociated body, i.e. a large cross-linking ratio.

\begin{table}
\begin{tabular}{l|SS|SSSS}
& {$G^\prime(\gamma\rightarrow 0)$} & {$\gamma_c$} & {$n$} & \multicolumn{2}{c}{$E_c$} & {$A$}\\
&	{\si{\pascal}} &  & & {$k_\mathrm{B}T$} & {\si{\kilo\joule/\mol}} & \\\hline&&&&\\[-10pt]
\ce{PPyr+I-}	& 8800	&	0.04	&	1.	&	2e-3\cellcolor{gray!25}	&	0.004\cellcolor{gray!25}	&	1.\\
\ce{PPyr+Br-}	& 9800 	& 	0.016	&	1.05&	3e-4\cellcolor{gray!25}	&	0.001\cellcolor{gray!25}	&	1.1\\
\ce{PPyr+Cl-}	& 270 	&	0.09	&	45	&	8e-3\cellcolor{gray!25}	&	0.02\cellcolor{gray!25}	&	7.\\
\ce{PIm+I-}	& 78	&	0.16	&	113	&	3e-2\cellcolor{gray!25}	&	0.063\cellcolor{gray!25}	&	9.6\\
\ce{PIm+Br-}	& 13	&	8.04	&	790	&	64	&	160.	&	3.4\cellcolor{gray!25}\\
\end{tabular}
\caption{Summary of rheological measurements and microscopic values deduced from the model. Gray background indicates incoherent values obtained either by (Eq.~\ref{eq:Ec}) or by (Eq.~\ref{eq:A}) outside of their respective validity domain.}
\label{tab:results}
\end{table}

@@

When the halide counterion is ``hard'', there is not enough charge dissociation to allow dissolution and gel formation. On the other hand of the spectrum outlined by HSAB theory, the hardest gels with the narrowest linear domain are obtained when a combination of ``hard'' cations and ``soft'' halides are used. This gives rise to maximal charge dissociation in water, facilitates charge pairing of the peripheral, iterative, cations with the terminal anion and thus afford a very high density of cross-links. In between, in the case of ``hard'' cations and not so hard halide (\ce{Cl-}) or in the ``soft-soft'' case, charge dissociation is limited, polymers associate in crocodile lines and cross-links are few. We thus obtain soft, highly deformable gels.

It is thus possible to obtain robust, highly tunable hydrogels with short, linear polymer chains and in the absence of any additive, which could find interesting applications especially in the context of biomaterials.

\begin{acknowledgement}
The authors are grateful for the financial support of Université de Lyon through the program ``Investissements d'Avenir'' ( ANR-1 1-IDEX-0007). S.M. and N.T. acknowledge funding from the European Research Council under the European Union's Seventh Framework Program (FP7/2007–2013)/ERC grant agreement No.258803.
\end{acknowledgement}

>>>>

Between two cross-links the crocodile chain can be considered as an entropic spring made of $N=n N_0$ monomers. The shear modulus of the gel is then given by\cite{Rubinstein1996}:
\begin{equation}
G = \frac{c}{N}k_\mathrm{B}T,
\label{eq:G}
\end{equation}
where $c$ is the monomer number density, $k_\mathrm{B}$ is the Boltzmann constant and $T$ the temperature. Because of the different weights of the pyrrolidinium and imidazolium moieties and of the different counter-ions, $c$ is not constant across our experiments done at constant weight fraction $w$ of polymer. This explains why the respective modulus or critical strains of \ce{PPyr+I-} (heavier, lower number density) and \ce{PPyr+Br-} are in reverse order with respect to the \textit{a priori} ranking of charge dissociation. Taking into account the molecular mass $M$ of each polymer and  knowing the density $d$ of the solvent, we obtain
\begin{equation}
n = \frac{\mathcal{N_A}}{M} w d \frac{k_\mathrm{B}T}{G}.
\label{eq:n}
\end{equation}

We find that the number $n$ of chains between cross-link point goes from 1 in \ce{PPyr+I-} and \ce{PPy+Br-} to 790 in \ce{PIm+Br-}, following the \textit{a priori} ranking of charge dissociation, see Table~\ref{tab:results}. This is consistent with a higher probability of attaching two or more heads on a highly dissociated body, i.e. a large cross-link ratio.

\begin{table}
\begin{tabular}{l|SS|SSSSS}
& {$G^\prime(\gamma\rightarrow 0)$} & {$\gamma_c$} & {$n$} & \multicolumn{2}{c}{$E_c$} & {$A$} & {$\sqrt{n}$}\\
&	{\si{\pascal}} &  & & {$k_\mathrm{B}T$} & {\si{\kilo\joule/\mol}} & \\\hline&&&&&\\[-10pt]
\ce{PPyr+I-}	& 8800	&	0.04	&	1.	&	2e-3\cellcolor{gray!25}	&	0.004\cellcolor{gray!25}	&	1.	&	1.\\
\ce{PPyr+Br-}	& 9800 	& 	0.016	&	1.05&	3e-4\cellcolor{gray!25}	&	0.001\cellcolor{gray!25}	&	1.1	&	1.\\
\ce{PPyr+Cl-}	& 270 	&	0.09	&	45	&	8e-3\cellcolor{gray!25}	&	0.02\cellcolor{gray!25}	&	7.	&	6.7\\
\ce{PIm+I-}	& 78	&	0.16	&	113	&	3e-2\cellcolor{gray!25}	&	0.063\cellcolor{gray!25}	&	9.6	&	10.6\\
\ce{PIm+Br-}	& 13	&	8.04	&	790	&	64	&	160.	&	3.4\cellcolor{gray!25}	&	28\\
\end{tabular}
\caption{Summary of rheological measurements and microscopic values deduced from the model. Gray background indicates inconsistent values obtained either by (Eq.~\ref{eq:Ec}) or by (Eq.~\ref{eq:A}) outside of their respective validity domain.}
\label{tab:results}
\end{table}

@@

When the halide counterion is ``hard'', there is not enough charge dissociation to allow dissolution and gel formation. On the other hand of the spectrum outlined by HSAB theory, the hardest gels with the narrowest linear domain are obtained when a combination of ``hard'' cations and ``soft'' halides are used. This gives rise to maximal charge dissociation in water, facilitates charge pairing of the peripheral, iterative cations with the terminal anion and thus affords a very high density of cross-links. In between, in the case of ``hard'' cations and not-so-hard halide (\ce{Cl-}) or in the ``soft-soft'' case, charge dissociation is limited, polymers associate in crocodile lines and cross-links are few. We thus obtain soft, highly deformable gels.

To conclude, our procedure yields robust, highly tunable hydrogels from short, linear polymer chains and in the absence of any additive, which could find interesting applications especially in the context of biomaterials and paves the way for future works on supra-molecular assembly.

\begin{acknowledgement}
The authors are grateful for the financial support of Université de Lyon through the program ``Investissements d'Avenir'' (ANR-1 1-IDEX-0007). S.M. and N.T. acknowledge funding from the European Research Council under the European Union's Seventh Framework Program (FP7/2007–2013)/ERC grant agreement No.258803.
\end{acknowledgement}
